---
title: "Sleep EDA - Final Project"
author: "Joshua LeVar"
date: "`r Sys.Date()`"
output: html_document
---
# A Statistical Analysis of Sleep Health Data
## Background

The purpose of this study is to use real-world data of sleep and health to determine how different sleep and health factors relate. I believe this to be interesting as it might reveal something interesting about the nature of the world. 
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
data = read.csv('./SleepStudy.csv', header=TRUE)
str(data)
```

## Research Questions:

In this project I will look to answer the following questions:

- Does gender affect sleep quantity?
- Does gender affect stress?
- Does alcohol use affect depression status?
- Does sleep quality affect cognition scores?**

## Methodology

### Data Description
**From the Textbook** 

> The data were obtained from a sample of students who did skills tests to measure cognitive function, completed a survey that asked many questions about attitudes and habits, and kept a sleep diary to record time and quality of sleep over a two-week period.

```{r}
dim(data)
```
There are 253 observations of 27 variables

**Variables of Interest**

- Gender (categorical): 1 = male, 0 = female
- AverageSleep (numeric): Average hours of sleep for all days
- Stress (categorical): Coded stress score: normal or high
- AlcoholUse (categorical): Self-reported: Abstain, Light, Moderate, or Heavy
- DepressionStatus (categorical): Coded depression score: normal, moderate, or severe
- PoorSleepQuality (numeric): Measure of sleep quality (higher values are poorer sleep)
- CognitionZscore (numeric): Z-score on a test of cognitive skills

### Data Analysis

To answer: *Does gender affect sleep quantity?*, I will use  a t-test to analyse the difference in means of sleep quantity by gender.

To answer: *Does gender affect stress?*, I will use a z-test to analyse the difference in proportions of stress status form males and females.

To answer: *Does alcohol use affect depression status?*, I will use a chi-square analysis by looking at the difference in proportions of depression status for all levels of alcohol use.

To answer: *Does sleep quality affect cognition scores?*, I will be using regression to analyse how a unit change in sleep quality causes a unit change in cognition scores**

### Hypotheses

#### t-test
- Null: The difference in the means is 0
- Alternative: The difference in the means is not 0

#### z-test
- Null: The difference in the proportions is 0
- Alternative: The difference in the proportions is not 0

#### chi-square
- Null: All proportions are equal
- Alternative: At least one proportion is not equal to the rest
 
#### regression
- Null: All regression coefficients are equal to 0
- Alternative: At least one coefficient is not 0

## Results
```{r}
table(data$Gender)
boxplot(data$AverageSleep ~ data$Gender)
```
```{r}
female = data[data$Gender == 0,]
male = data[data$Gender == 1,]
c(summary(female$AverageSleep), setNames(sd(female$AverageSleep), "SD"))
```

```{r}
c(summary(male$AverageSleep), setNames(sd(male$AverageSleep), "SD"))
```
```{r}
t.test(data$AverageSleep ~ data$Gender, alternative="two.sided", conf.level=0.95)
```
## Discussion



