---
title: "Sleep EDA - Final Project"
author: "Joshua LeVar"
date: "`r Sys.Date()`"
output: html_document
---
# A Statistical Analysis of Sleep Health Data
## Background

The purpose of this study is to use real-world data of sleep and health to determine how different sleep and health factors relate. I believe this to be interesting as it might reveal something interesting about how our decisions about sleep and life affect each other.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
data = read.csv('./SleepStudy.csv', header=TRUE)
str(data)
```

## Research Questions:

In this project I will look to answer the following questions:

- Does gender affect sleep quantity?
- Does gender affect stress?
- Does alcohol use affect depression status?
- How do cognition scores depend on sleep quality, sleep quantity, and happiness?

## Methodology

### Data Description
**From the Textbook** 

> The data were obtained from a sample of students who did skills tests to measure cognitive function, completed a survey that asked many questions about attitudes and habits, and kept a sleep diary to record time and quality of sleep over a two-week period.

```{r}
dim(data)
```
There are 253 observations of 27 variables

**Variables of Interest**

- Gender (categorical): 1 = male, 0 = female
- AverageSleep (numeric): Average hours of sleep for all days
- Stress (categorical): Coded stress score: normal or high
- AlcoholUse (categorical): Self-reported: Abstain, Light, Moderate, or Heavy
- DepressionStatus (categorical): Coded depression score: normal, moderate, or severe
- PoorSleepQuality (numeric): Measure of sleep quality (higher values are poorer sleep)
- CognitionZscore (numeric): Z-score on a test of cognitive skills
- Happiness (numeric): Measure of degree of happiness

### Data Analysis

To answer: *Does gender affect sleep quantity?*, I will use  a t-test to analyse the difference in means of sleep quantity by gender.

To answer: *Does gender affect stress?*, I will use a z-test to analyse the difference in proportions of stress status form males and females.

To answer: *Does alcohol use affect depression status?*, I will use a chi-square analysis by looking at the difference in proportions of depression status for all levels of alcohol use.

To answer: *How do cognition scores depend on sleep quality, sleep quantity, and happiness?*, I will be using multiple regression to analyse what variables, if any, have a statistically significant affect on cognition scores

### Hypotheses

#### T-Test
- Null: The difference in the means is 0
- Alternative: The difference in the means is not 0

#### Z-Test
- Null: The difference in the proportions is 0
- Alternative: The difference in the proportions is not 0

#### Chi-Squared
- Null: All proportions are equal
- Alternative: At least one proportion is not equal to the rest
 
#### Multiple Regression
- Null: All regression coefficients are equal to 0
- Alternative: At least one coefficient is not 0

## Results
### Does gender affect sleep quantity?
```{r}
boxplot(data$AverageSleep ~ data$Gender)
```
```{r}
female = data[data$Gender == 0,]
male = data[data$Gender == 1,]
c(summary(female$AverageSleep), setNames(sd(female$AverageSleep), "SD"))
c(summary(male$AverageSleep), setNames(sd(male$AverageSleep), "SD"))
t.test(data$AverageSleep ~ data$Gender, alternative="two.sided", conf.level=0.95)
```

#### Conclusion
The t-test results show no statistically significant difference between the
average sleep times between Male and Female (p-value 0.5585 > 0.05)

### Does gender affect stress?
```{r}
table(data$Gender, data$Stress)
```
```{r}
proportions = setNames(c(mean(female$Stress == "high"), mean(male$Stress == "high")), c("Female", "Male"))
counts = setNames(c(sum(female$Stress == "high"), sum(male$Stress == "high")), c("Female", "Male"))
proportions
```

```{r}
barplot(height = proportions,
        ylim = c(0, 1), col = "skyblue",
        ylab = "Proportion", main = "Proportion of High Stress by Gender")
```

```{r}
prop.test(x=counts, n=c(nrow(female), nrow(male)), alternative="two.sided", conf.level=0.95)
```
#### Conclusion
The z-test results show no statistically significant difference between the
proportion of high stress individuals between Male and Female (p-value 0.2081 > 0.05)

### Does alcohol use affect depression status?
```{r}
alcoholDepressionTable = table(data$AlcoholUse, data$DepressionStatus)
alcoholDepressionTable
```
```{r}
df = as.data.frame(as.table(alcoholDepressionTable))
colnames(df) = c("AlcoholUse", "DepressionStatus", "Count")
df$AlcoholUse = factor(df$AlcoholUse, levels = c("Abstain", "Light", "Moderate", "Heavy"))
df$DepressionStatus = factor(df$DepressionStatus, levels = c("severe", "moderate", "normal"))

ggplot(df, aes(x = AlcoholUse, y = Count, fill = DepressionStatus)) +
  geom_bar(stat = "identity", position = "fill") +
  labs(y = "Proportion", title = "Proportion of Depression Status by Alcohol Use") +
  scale_fill_manual(values = c(
  "normal" = "#a6d9e7",
  "moderate" = "#f9c74f",
  "severe" = "#f94144")) +
  theme_minimal()
``` 
```{r}
chisq.test(alcoholDepressionTable)
```
#### Conclusion
The Chi-squared test results show no statistically significant difference between the
proportions of depression status between different categories of alcohol use 
(p-value 0.5629 > 0.05)

### How do cognition scores depend on sleep quality, sleep quantity, and happiness?

```{r}
model <- lm(CognitionZscore ~ PoorSleepQuality + AverageSleep + Happiness, data = data)

summary(model)
anova(model)
par(mfrow = c(2, 2)) 
plot(model)
```

#### Conclusion
The multiple regression model shows no statistically significant effect of sleep quality,
sleep quantity, or happiness on cognition scores
(all p-values > 0.05)

## Discussion
### Meaning
I was quite surprised to see that none of my tests gave any statistically significant result.
I was hoping to see at least some evidence of a factor that affects or is affected my sleep. 
In short, we were not able to show any:

- Effect of gender on sleep quantity
- Effect of gender on stress
- Effect of alcohol on depression status
- Effect of sleep quality, sleep quantity, or happiness on cognition scores

### Limitations
I believe that sample size was not an issue here. Since none of our results are conclusive,
I don't see much reason to be concerned about bias or confounders. It's not obvious to me
that such confounders exist that would hide the statistical assocations that we are looking at
in this analysis.


